Spatiotemporal models of species biomass split by maturity class

### 2019-09-10 analysis/VOCC/1-split-biomass-by maturity.Rmd
give splitting my maturity it's own rmd

# species <- "Walleye Pollock"
# species <- "Pacific Cod"
# species <- "Pacific Ocean Perch"

# species <- "Silvergray Rockfish"
# species <- "Quillback Rockfish"
# species <- "Yelloweye Rockfish"
# species <- "Bocaccio"
# species <- "Sharpchin Rockfish"
# species <- "Splitnose Rockfish"
# species <- "Canary Rockfish"
# species <- "Rougheye/Blackspotted Rockfish Complex"
 species <- "Redbanded Rockfish"
# species <- "Greenstriped Rockfish"

# species <- "Longspine Thornyhead"
# species <- "Shortspine Thornyhead",
# species <- "Sablefish"
# species <- "Lingcod"

# species <- "North Pacific Spiny Dogfish" # note: using all data for maturity thresholds
# species <- "Longnose Skate"
 species <- "Spotted Ratfish"

```

Region 
```{r}
  region <- "All synoptic surveys"
 # region <- "Both odd year surveys"
 # region <- "West Coast Vancouver Island"
 # region <- "West Coast Haida Gwaii"
```

Choose model details
```{r}
priors <- FALSE
# priors <- TRUE
# covariates <- "+muddy+any_rock"
covariates <- ""
covs <- gsub("\\+", "-", "+muddy+any_rock")
#covs <- gsub("\\+", "-", covariates)
# covariates <- "+muddy+mixed+rocky"
# covariates <- "+mixed+rocky"
# covariates <- "+trawled+muddy+rocky+mixed"
# covariates <- "+trawled+mixed+rocky"
# covariates <- "+trawled+mixed"
```

@@

  if(is.null(maturity$maturity)) {
  maturity <- split_catch_maturity(survey_sets, fish, bath,
    survey = c("SYN QCS","SYN HS","SYN WCVI","SYN WCHG"),
    years = years,
    cutoff_quantile = c(.9995),
    plot = TRUE
  )
  maturity$data <- filter(maturity$data, ssid == model_ssid) 
  }

@@

# Build spatiotemporal density model
Load and filter data 
```{r}
biomass <- readRDS(paste0(
    "data/", spp, "/data-by-maturity-", spp, "-", ssid_string, ".rds"
    ))
  covars <- readRDS("data/event-covariates.rds")
  data <- dplyr::left_join(biomass, covars)

  # scale predictors before filtering to ensure mean and SD are global
  data <- data %>% mutate(raw_depth = depth, depth = log(raw_depth)) 
  data <- gfranges::scale_predictors(data, predictors = c(quo(depth)))
       # c(quo(log_depth), quo(mixed), quo(muddy), quo(sandy), quo(rocky), quo(any_rock)))
  data <- data %>% mutate(depth = raw_depth) %>% filter(ssid %in% model_ssid)
```

Run climate independent sdmTMB model
```{r}

  if (region == "Both odd year surveys") {
    spde <- sdmTMB::make_spde(data$X, data$Y, n_knots = 400)
  }

  if (region == "West Coast Vancouver Island") {
    spde <- sdmTMB::make_spde(data$X, data$Y, n_knots = 200)
  }

  if (region == "West Coast Haida Gwaii") {
    spde <- sdmTMB::make_spde(data$X, data$Y, n_knots = 200)
  }

  if (region == "All synoptic surveys") {
    spde <- sdmTMB::make_spde(data$X, data$Y, n_knots = 500)
  }

  sdmTMB::plot_spde(spde)

  if (any(names(data) == "adult_density")) {

    adult_formula <- as.formula(paste(
      "adult_density ~ 0 + as.factor(year)", covariates, ""
      ))

    starttime1 <- Sys.time()
    adult_biomass <- sdmTMB::sdmTMB(data,
      adult_formula,
      time_varying = ~ 0 + depth_scaled + depth_scaled2 + trawled,
      time = "year", spde = spde,
      family = tweedie(link = "log"),
      ar1_fields = TRUE,
      include_spatial = TRUE,
      enable_priors = priors,
      control = sdmTMBcontrol(step.min = 0.01, step.max = 1),
      silent = FALSE
    )
    endtime1 <- Sys.time()
    time1 <- round(starttime1 - endtime1)

    saveRDS(adult_biomass, 
      file = paste0("data/", spp, 
        "/mod-mat-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
        ))

    imm_formula <- as.formula(paste(
      "imm_density ~ 0 + as.factor(year)", covariates, ""
      ))

    starttime2 <- Sys.time()
    imm_biomass <- sdmTMB::sdmTMB(data, 
      imm_formula,
      time_varying = ~ 0 + depth_scaled + depth_scaled2 + trawled,
      time = "year", spde = spde,
      family = tweedie(link = "log"),
      ar1_fields = FALSE,
      include_spatial = TRUE,
      enable_priors = priors,
      control = sdmTMBcontrol(step.min = 0.01, step.max = 1),
      silent = FALSE
    )
    endtime2 <- Sys.time()
    time2 <- round(starttime2 - endtime2)

    saveRDS(imm_biomass, 
     file = paste0("data/", spp, 
       "/mod-imm-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
       ))

  } else {

    starttime3 <- Sys.time()
    dens_formula <- as.formula(paste("density ~ 0 + as.factor(year)", covariates, ""))

    total_biomass <- sdmTMB::sdmTMB(data,
      dens_formula,
      time_varying = ~ 0 + depth_scaled + depth_scaled2 + trawled,
      time = "year", spde = spde,
      family = tweedie(link = "log"),
      ar1_fields = TRUE,
      include_spatial = TRUE,
      enable_priors = priors,
      control = sdmTMBcontrol(step.min = 0.01, step.max = 1),
      silent = FALSE
    )
    endtime3 <- Sys.time()
    time3 <- round(starttime3 - endtime3)

    saveRDS(total_biomass, file = paste0("data/", spp, 
      "/model-total-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
      ))
  }
```

Make prediction grids for all surveys and years, if not done before.
```{r, message=FALSE, warning=FALSE}
try(nd_all <- readRDS("data/nd_all_synoptic.rds"))

if (!exists("nd_all")) {

  # choose base year(s) to create grid from
  dummy_year <- c(2005, 2006)

  bath <- readRDS("data/bathymetry-data")
  #dat <- bath$data %>% scale_survey_predictors()
  dat <- bath$data %>% mutate(raw_depth = depth, depth = log(raw_depth)) 
  dat <- gfranges::scale_predictors(dat, predictors = c(quo(depth)))

  # create grids for each ssid separately
  ssid <- 1
  survey_abbrev <- "SYN QCS"
  nd_1 <- spatiotemporal_grid(dat, ssid, survey_abbrev, dummy_year)
  unique(nd_1$year)

  ssid <- 3
  survey_abbrev <- "SYN HS"
  nd_3 <- spatiotemporal_grid(dat, ssid, survey_abbrev, dummy_year)
  unique(nd_3$year)

  ssid <- 4
  survey_abbrev <- "SYN WCVI"
  nd_4 <- spatiotemporal_grid(dat, ssid, survey_abbrev, dummy_year)
  unique(nd_4$year)

  ssid <- 16
  survey_abbrev <- "SYN WCHG"
  nd_16 <- spatiotemporal_grid(dat, ssid, survey_abbrev, dummy_year)
  unique(nd_16$year)

  nd_all <- rbind(nd_1, nd_3, nd_4, nd_16) %>% 
    mutate(year = as.numeric(year))

  newcov <- readRDS("data/new_covariates.rds") %>% dplyr::select(X, Y, trawled, muddy, rocky, mixed, any_rock)
  nd_all <- left_join(nd_all, newcov, by=c("X", "Y"))

  # nd <- nd %>% dplyr::mutate(shallow = ifelse(depth > 35, 0, 1))

  saveRDS(nd_all, file = paste0("data/nd_all_synoptic.rds"))
}
```

Save predictions for full spatial grid

```{r}
#covs <- gsub("\\+", "-", covariates)
fish <- readRDS(paste0("raw/bio-data-", spp, ""))

rm(adult_predictions)
rm(imm_predictions)
rm(predictions)

if (length(unique(fish$maturity_code)) > 2) {

    nd_all <- readRDS(paste0("data/nd_all_synoptic.rds"))
    adult_biomass <- readRDS(paste0("data/", spp, 
      "/mod-mat-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
    ))

    nd <- nd_all %>% 
      filter(ssid %in% model_ssid) %>% 
      filter(year %in% unique(adult_biomass$data$year))
    nd <- na.omit(nd)
    nd$year <- as.integer(nd$year)

    adult_predictions <- predict(adult_biomass, newdata = nd)
    saveRDS(adult_predictions, file = paste0("data/", spp, 
      "/predictions-", spp, covs, "-", ssid_string, "-mature-biomass-prior-", priors, ".rds"
    ))

    imm_biomass <- readRDS(paste0("data/", spp, 
      "/mod-imm-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
    ))
    imm_predictions <- predict(imm_biomass, newdata = nd)
    saveRDS(imm_predictions, file = paste0("data/", spp, 
      "/predictions-", spp, covs, "-", ssid_string, "-imm-biomass-prior-", priors, ".rds"
    ))
  } else {
    rm(adult_predictions)
    rm(imm_predictions)

    if (!exists(predictions)) {
      nd_all <- readRDS(paste0("data/nd_all_synoptic.rds"))
      nd <- nd_all %>% filter(ssid %in% model_ssid)
      nd <- na.omit(nd)
      nd$year <- as.integer(nd$year)

      total_biomass <- readRDS(paste0("data/", spp, 
        "/model-total-biomass-", spp, covs, "-", ssid_string, "-prior-", priors, ".rds"
      ))
      predictions <- predict(total_biomass, newdata = nd)
      saveRDS(predictions, file = paste0("data/", spp, 
        "/predictions-", spp, covs, "-", ssid_string, "-total-biomass-prior-", priors, ".rds"
      ))
    }
  }
```

>>>>

# species <- "Silvergray Rockfish"
# species <- "Quillback Rockfish" # summer-birthing
species <- "Yelloweye Rockfish" # summer-birthing, overfished, 
# species <- "Bocaccio" # winter-birthing, overfished
# species <- "Sharpchin Rockfish"
# species <- "Splitnose Rockfish"

# species <- "Rougheye/Blackspotted Rockfish Complex"
# species <- "Redbanded Rockfish"
# species <- "Greenstriped Rockfish"
# # species <- "Copper Rockfish" # small sample
# species <- "Darkblotched Rockfish"
# # species <- "Shortbelly Rockfish" # small sample

# species <- "Pacific Ocean Perch" # schooling
# species <- "Widow Rockfish" # schooling
# species <- "Yellowtail Rockfish" # schooling
# species <- "Canary Rockfish" # schooling, winter-birthing

# species <- "Longspine Thornyhead"
# species <- "Shortspine Thornyhead"

# species <- "Walleye Pollock"
# species <- "Pacific Cod"
# species <- "Sablefish"
# species <- "Lingcod"
# species <- "Pacific Hake"

# species <- "North Pacific Spiny Dogfish" # note: using all data for maturity thresholds
# species <- "Longnose Skate"
# species <- "Big Skate"
# species <- "Spotted Ratfish"
# # species <- "Sandpaper Skate" # small sample
# # species <- "Brown Cat Shark" # small sample
```

Region 
```{r}
# region <- "All synoptic surveys"
 region <- "Both odd year surveys"
 # region <- "West Coast Vancouver Island"
 # region <- "West Coast Haida Gwaii"
```

@@

  if(is.null(maturity$maturity)) {
  maturity <- split_catch_maturity(survey_sets, fish, bath,
    survey = c("SYN QCS","SYN HS","SYN WCVI","SYN WCHG"),
    years = years,
    cutoff_quantile = c(.9995),
    plot = TRUE
  )
  }
  maturity$data <- filter(maturity$data, ssid == model_ssid) 
  # %>% select(-adult_density, -imm_density) # if maturity model is deemed unuseful

@@
